---
title: "Lab 6 Key - Spatial data viz, kriging, rasters intro"
author: "Allison Horst"
date: "2/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Attach packages:
```{r}
library(tidyverse)
library(here)
library(janitor)
library(raster)
library(sf)
library(tmap)
library(gstat) # variogram
library(plotKML) # convert points back to raster
```

## Grand Canyon DEM 

####  Read in some raster data (GeoTIFF DEM)

Grand Canyon GeoTIFF from USGS: https://pubs.usgs.gov/ds/121/grand/grand.html

```{r}
# Read in the data with raster::raster()
gc_dem <- raster(here("data","gc_dem.tif"))
```

#### Go exploring (basic plot, check CRS, projection, bounds)
```{r}
# Look at it with base plot()
plot(gc_dem)

# Check CRS & bounds:
gc_dem@crs # Shows CRS: WGS84
gc_dem@extent # Shows extent (bounds)...notice that these seem odd (not units)

wgs84 = "+proj=longlat +datum=WGS84 +ellps=WGS84 +no_defs" # Just have this ready to copy/paste

# Reproject
gc_reproj = projectRaster(gc_dem, crs = wgs84, method = "bilinear")

# Then check: aha, now degrees we're used to
gc_reproj@extent

```

#### Crop the raster to a smaller area: 
```{r}
bounds <- as(extent(-112.4, -112.0, 36.1, 36.3), 'SpatialPolygons')

# Make the projection for "bounds" the same as for "gc_reproj":
crs(bounds) <- crs(gc_reproj)

# Then crop gc_reproj by the new bounds polygon: 
gc_crop <- crop(gc_reproj, bounds)

# Look at it:
plot(gc_crop)
```

Want to resample? Use `raster::aggregate()` to create lower res (larger cell) rasters.

See ?aggregrate (default is mean, `fact` is number of cells in each direction; can set 2 if wanting x/y to differ for aggregation)
```{r}
# Aggregate: 
gc_agg <- aggregate(gc_crop, fact = 10)

# Then look at it: 
plot(gc_agg)
```

### Get safe and sound back in ggplot:
```{r}
# First, convert to a data frame:
gc_df <- as.data.frame(gc_agg, xy = TRUE)

# That `xy = TRUE` is important: retains the lat/lon information! 

# View(gc_df)

ggplot(data = gc_df, aes(x = x, y = y)) +
  geom_raster(aes(fill = gc_dem)) +
  coord_quickmap() +
  theme_minimal() +
  scale_fill_gradientn(colors = c("purple",
                                  "magenta",
                                  "orange",
                                  "yellow", 
                                  "white")
                       )
```

Let's say we know that in this region, a certain species will only grow between 1000 and 1500 ft elevation. Create a subset of gc_crop that includes that habitat:
```{r}
# First, make a copy
gc_hab <- gc_crop

# Set any cells outside of 1000 - 1500 to NA
gc_hab[gc_hab > 1500 | gc_hab < 1000] <- NA

# Plot.
plot(gc_hab)

# Cool! 
```

Let's make a bit nicer map with `tmap`: 
```{r}
tmap_mode("view") # Set to interactive viewing

# Make tmap: 
tm_shape(gc_hab) +
  tm_raster(legend.show = FALSE, palette = "plasma")

# In console, run tmaptools::palette_explorer() to view a Shiny app with other palettes!
```

## Kriging Kansas Rain

#### Read in the KS counties shapefile with `read_sf`:
```{r}
ks_counties <- read_sf(here("data",
                            "ks_counties",
                            "ks_counties_shapefile.shp"))

# View(ks_counties)
```

#### Do some exploring:
```{r}
# Base plot
plot(ks_counties)

# Check CRS:
st_crs(ks_counties) # hmmmm none...guess we should set one!

# Set to EPSG 4326 (WGS84 datum):
st_crs(ks_counties) <- 4326

# Now check again:
st_crs(ks_counties)

# And replot:
plot(ks_counties)

# That looks more like Kansas. 
```

#### Now, read in the KS rainfall data (this was from a single rainy day in 2017): 
```{r}
ks_rain <- read_csv(here("data","ks_rain.csv")) %>% 
  clean_names()
```

But currently, R has no idea that these are spatial points. We'll convert it using `sf::st_as_sf()`:

```{r}
ks_sf  <-  st_as_sf(ks_rain, coords = c("lon", "lat"), 
                 crs = 4326)

# View(ks_sf) (rainfall = amt)
plot(ks_sf)

# Or in ggplot:
ggplot() +
  geom_sf(data = ks_counties) +
  geom_sf(data = ks_sf, aes(color = amt,
              size = amt),
          show.legend = FALSE) +
  coord_sf() +
  scale_color_gradient(low = "yellow", high = "red") +
  theme_void()
```

Now let's try to predict rainfall all over Kansas based on those rainfall values. 

For kriging, we're gonna switch over to `sp` functions. Which means we need to get our data to 'Spatial' format, instead of `sf`. Eventually, I think these will all be merged...

#### Make the rainfall data a Spatial Points data frame (so we can use gstat/sp functions: 
```{r}
ks_sp  <- as_Spatial(ks_sf)
# class(ks_sp)
```

#### Make a grid that we'll krige over:
```{r}
# bbox(ks_sp) to check bounding box of the spatial points
lat <- seq(37, 40, length.out = 200)
long <- seq(-94.6,-102, length.out = 200)

# Then make it into a grid: 
grid <- expand.grid(lon = long, lat = lat)
grid_sf <- st_as_sf(grid, coords = c("lon","lat"), crs = 4326)
grid_sp <- as_Spatial(grid_sf)

# Check out your amazing kriging-ready grid:
plot(grid_sp)
```

#### Make a variogram from the rainfall data: 
```{r}

# Create the variogram:
ks_vgm <- variogram(amt ~ 1, data = ks_sp)

# Look at it: 
plot(ks_vgm)

# Fit the variogram model using reasonable estimates for nugget, sill and range:
ks_vgm_fit <- fit.variogram(ks_vgm, model = vgm(nugget = 0.2, psill = 0.8, model = "Sph", range = 200))

# Plot them both together
plot(ks_vgm, ks_vgm_fit) # Cool! So what are the values

# Just FYI: there are other models (Gaussian, Exponential) - how do those line up? 
ks_vgm_gau <- fit.variogram(ks_vgm, model = vgm(nugget = 0.2, psill = 0.8, model = "Gau", range = 200))

plot(ks_vgm, ks_vgm_gau)

# You can check the sum of squares of residuals for each: 
attr(ks_vgm_fit, 'SSErr') # 0.00214 (and could compare to other models...)

# We'll stick with the Spherical model: 
ks_vgm_fit # Nugget = 0.102, sill = 0.954, range = 235
```

Now, kriging! 
```{r}

ks_krige <- krige(amt ~ 1, ks_sp, grid_sp, model=ks_vgm_fit)

```

And visualize it: 
```{r}

ks_krige_df <- as.data.frame(ks_krige) # View it after this to show output

# Rename things to make it a little nicer
ks_krige_2 <- ks_krige_df %>% 
  rename(lon = coords.x1, lat = coords.x2, predicted = var1.pred, err = var1.var)

# Make this into spatial data again
rain_predicted  <-  st_as_sf(ks_krige_2, coords = c("lon", "lat"), crs = 4326)

```

Get state outlines & filter to just get KS:
```{r}
# Get Kansas outline to crop: 
ks <- read_sf(dsn = here("data","states"), 
              layer = "cb_2017_us_state_20m") %>% 
  dplyr::select(NAME) %>% 
  filter(NAME == "Kansas") %>% 
  st_transform(crs = 4326)

# Crop the rainfall data
rain_cropped <- st_intersection(rain_predicted, ks)

# Initial plot
plot(rain_cropped) # But this is points

# So is this (cheating...)
# tmap: 
  tm_shape(rain_cropped) +
  tm_dots("predicted", size = 0.05) +
  tm_shape(ks_counties) +
  tm_borders() +
    tm_layout(legend.bg.color = "white", legend.position = c("left","bottom"))

```

Extra...converting sf points to Spatial points to raster (with plotKML package): 

```{r}

# Convert sf object to spatial points
rain_pts <- as_Spatial(rain_cropped)
class(rain_pts)

# Rasterize spatial points, make class 'Raster'
rain_raster <- vect2rast(rain_pts, cell.size = 0.01)
rain_raster2 <- raster(rain_raster)

# Need to aggregate so it's not just tiny cells (and white space)
rain_raster_agg <- raster::aggregate(rain_raster2, fact = 5, fun = mean)

tm_shape(rain_raster_agg) +
  tm_raster() +
  tm_shape(ks_counties) +
  tm_borders() +
    tm_layout(legend.bg.color = "white", legend.position = c("left","bottom"))

```

